Autoregressive Crosslagged Panel Models

### 2024-03-06 _posts/2_Examples_ResidualizedChangeScores.qmd
revised manuscript docs

---
title: "Difference Scores: Simple Regression"
author: "Biostatistics Working Group"

reader-mode: true
---

# Simple Regression {.unnumbered}

@@

This example assesses whether growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1) differs significantly based on handedness, using height ("anthroheightcalc") as a representative metric of growth. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 height measurements for each participant; 2) a simple regression analysis is used to test whether sex (boy, girl) predicts the average difference value in participants height from baseline to the 1-Year follow-up. Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between difference scores and sex. The ensuing analysis and interpretations are detailed in the subsequent sections.

@@

# rm(list = ls())

@@

::: panel-tabset
### Build Model: Step 1 {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to compute a difference score by subtracting each participant’s height at T1 from their height at T0. Relevant summary statistics are also provided.

**STEP 1: Compute Difference Score**
```{r build-model-regression-1}
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| class.source: 'fold-hide'

# Define the function to compute difference scores for a given variable and provide a summary
# Function to compute difference scores for a given variable and provide a summary
compute_difference_and_summary <- function(df, variable_name) {
    # Define the event names of interest
    baseline_event <- "baseline_year_1_arm_1"
    followup_event <- "1_year_follow_up_y_arm_1"

    # Compute the difference between Baseline and Year 1 data for the given variable
    diff_data <- df %>%
        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
        spread(eventname, variable_name) %>% # Convert data from long to wide format
        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
        drop_na(diff) # Exclude rows with NA in the computed difference

    # Summarize the computed difference scores
    diff_summary <- summary(diff_data$diff)

    # Return the difference data and its summary
    list(data = diff_data, summary = diff_summary)
}

# List of variables for which difference scores are to be computed
variables_of_interest <- c("anthroheightcalc")

# Compute and store difference scores and summaries for each variable in a list
difference_and_summary_list <- lapply(variables_of_interest, function(var) {
    compute_difference_and_summary(df, var)
})

# Extract the difference data for the 'anthroheightcalc' variable
height_diff_data <- difference_and_summary_list[[1]]$data

# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
df <- left_join(df, height_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")

@@

**Descriptive Statistics for the Difference Score**
```{r model-checking}
# Compute statistical summaries for the difference score variable
lapply(difference_and_summary_list, function(item) {
    print(item$summary)
})
```

This summary of the difference score variable indicates that the differences range from a decrease of `r min(height_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(height_diff_data$diff, na.rm = TRUE)` units. The median difference is `r median(height_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(height_diff_data$diff, digits = 2, na.rm = TRUE)` units. There are `r sum(is.na(height_diff_data$diff))` missing values in this difference score dataset.

**Descriptive Statistics for the Height Variable**
```{r}
## Summary statistics
# Compute summary statistics for Height by eventname
summary <- df %>%
    group_by(eventname) %>%
    get_summary_stats(anthroheightcalc, type = "mean_sd")

data.frame(summary)
```

The summary statistics provide insights into the Height variable across different `eventname` categories. At baseline, the average height is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`. Over the years, there’s a noticeable increase in average height: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in height measurements across these time points.

@@

### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}

::: blue
A simple regression analyses is conducted to examine whether a grouping variable (sex) significantly predicts the difference score value (indicating significant group differences in the average difference score).

**STEP 2: Conduct regression on Difference Score**
```{r build-model-regression-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
merged_data <- height_diff_data %>%
  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")

# Ensure 'sex' is a factor
merged_data$sex <- as.factor(merged_data$sex)

# Run the regression
model <- lm(diff ~ sex, data = merged_data)

# Get the summary of the regression model
model_summary <- summary(model)

model_summary

@@

This regression analysis evaluates whether sex (boy, girl) predicts the average difference in participants' height from baseline to the 1-Year follow-up. The output from our model provides:

i. an F-statistic of `r model_summary$fstatistic[1]`; 
ii. degrees of freedom of `r model_summary$df[2]`; 
iii. a parameter estimate of `r model_summary$coefficients["sexMale or other", "Estimate"]`; 
iv. standard error of `r model_summary$coefficients["sexMale or other", "Std. Error"]`; 
v. p-value of `r model_summary$coefficients["sexMale or other", "Pr(>|t|)"]`.

Compared to boys (the reference group), girls have an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units. This effect was marginally significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`. Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was not statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.

@@

### Model Plots

::: blue
```{r violin-plots}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Visualize the difference scores across different levels of sex

# Create a violin plot to show the distribution of difference scores by sex
# Jittered points are added to provide a more granular view of individual observations
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
merged_data <- height_diff_data %>%
  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")

# Plotting using ggplot2
ggplot(merged_data, aes(x = sex, y = diff, fill = sex)) +
  geom_violin() +
  geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "Difference Scores by Sex",
    x = "Sex",
    y = "Difference Score"
  ) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

@@

The violin plot visualizes the distribution of difference scores in children's heights by their sex: Right, Left, and Mixed. Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points. Superimposed jittered points offer a granular view of individual observations. From the plot, it appears that the distributions of height differences across the three handedness categories are somewhat similar, though there might be subtle variations in median and spread.

@@

## Wrapping Up

@@

::: blue
Differences in children's heights between baseline and a subsequent 1-year follow-up, we sought to understand the potential influence of sex on this difference. Descriptive statistics revealed that height differences ranged from a decrease of `r min(merged_data$diff, na.rm = TRUE)` units to an increase of `r max(merged_data$diff, na.rm = TRUE)` units, with a median difference of `r median(merged_data$diff, na.rm = TRUE)` units and an average difference of approximately `r mean(merged_data$diff, na.rm = TRUE)` units. Subsequent regression analysis indicated that, when compared to boys, girls had an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units, though this effect was marginally significant (p = `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`). Complementary to these findings, violin plots visually underscored the subtle variations in height differences across sex categories, suggesting broadly similar distributions but with nuanced variations in central tendency and spread. In conclusion, while sex exhibited a potential influence on height difference over the year, the observed effect was notably minor.

>>>>

---
title: "Residualized Change Scores"
author: "Biostatistics Working Group"
---

# Mulitple Regression {.unnumbered}

@@

This example assesses whether height, measured at the 1-Year follow-up (T1) in a subsample of ABCD participants, is predicted by weight and height measured at baseline (T0). A visual inspection is further conducted via a scatterplot to graphically represent the relationship between height at Year_1 and weight at baseline, controlling for height at baseline. The ensuing analysis and interpretations are detailed in the subsequent sections.

@@

@@

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to conduct a multiple regression analysis by subtracting each participant's height at T1 from their height at T0. Relevant summary statistics are also provided.

**Compute Multiple Regression Model**
```{r build-model-resid-regression-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Split data using the correct eventname values
baseline_data <- df %>% 
  filter(eventname == "baseline_year_1_arm_1") %>% 
  select(src_subject_id, Height_baseline = anthroheightcalc, Weight_baseline = anthroweightcalc)

followup_data <- df %>% 
  filter(eventname == "1_year_follow_up_y_arm_1") %>% 
  select(src_subject_id, Height_followup = anthroheightcalc)

# Merge baseline and follow-up data
merged_data <- baseline_data %>%
  left_join(followup_data, by = "src_subject_id")

# Check the first few rows of the merged data
#head(merged_data)

# Fit the regression model excluding NA values
model <- lm(Height_followup ~ Height_baseline + Weight_baseline, data = merged_data, na.action = na.exclude)

@@

@@

### Model Summary {.tabset .tabset-fade .tabset-pills}

::: blue

A plot to show xxxxx.
```{r model-checking}
model_summary <- summary(model)
print(model_summary)

par(mfrow = c(2, 2))  # Set up a 2x2 plotting area
plot(model)  # Generate diagnostic plots

@@

This regression analysis evaluates how baseline height and weight predict children's height at a 1-year follow-up. The output from our model provides several key statistics:

i. an F-statistic of `r model_summary$fstatistic[1]`;
ii. degrees of freedom of `r model_summary$df[2]`;
iii. parameter estimates for baseline height and weight;
iv. standard errors for these estimates; and
v. p-values for the significance of these estimates.

For every one unit increase in baseline height, there's an associated increase of approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units in the follow-up height, and this effect was statistically significant with a p-value of `r model_summary$coefficients["Height_baseline", "Pr(>|t|)"]`. Similarly, for every one unit increase in baseline weight, there's an associated increase of approximately `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units in the follow-up height, which is also statistically significant with a p-value of `r model_summary$coefficients["Weight_baseline", "Pr(>|t|)"]`. Overall, this model explains a substantial portion of the variance in follow-up height, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 4)`. The overall model is highly significant with a p-value less than `r model_summary$coefficients[1,4]`, indicating that the predictors, collectively, have a significant relationship with the dependent variable.

@@

### Model Plots  {.tabset .tabset-fade .tabset-pills}

::: blue
```{r scatterplot}

# Scatterplot with regression line for Weight_baseline
ggplot(merged_data, aes(x = Weight_baseline, y = Height_followup)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") + 
  labs(title = "Relationship between Baseline Weight and Follow-up Height",
       x = "Baseline Weight",
       y = "Follow-up Height") +
  theme_minimal()

@@

The scatterplot visually depicts the relationship between children's weights at baseline (`r names(merged_data)[3]`) and their heights at Year_1 (`r names(merged_data)[4]`). Each point on the plot represents a child, with their baseline weight plotted on the x-axis and their Year_1 height on the y-axis. The clear positive linear trend, as illustrated by the black regression line, indicates that children who had higher weights at baseline generally had higher heights at Year_1. While further statistical analyses can quantify the strength and direction of this relationship, visually, the data suggests a pronounced positive association between baseline weight and follow-up height.

@@

## Wrapping Up
<div class = "blue">

@@

::: blue

The regression analysis was conducted to predict children's height at the 1-Year follow-up using their baseline height (`r names(merged_data)[2]`) and weight (`r names(merged_data)[3]`). Both predictors were statistically significant. Specifically, for every one-unit increase in baseline height, the height at Year_1 increased by approximately `r round(model_summary$coefficients["Height_baseline", "Estimate"], 3)` units, while holding weight constant. Similarly, for every one-unit increase in baseline weight, the height at Year_1 increased by about `r round(model_summary$coefficients["Weight_baseline", "Estimate"], 4)` units, while holding baseline height constant. The overall model explained a substantial `r round(model_summary$adj.r.squared * 100, 2)`% of the variance in height at the 1-Year follow-up, as indicated by the adjusted R-squared value.

### 2024-03-06 _posts/3a_Examples_LinearMixedModels.qmd
revised manuscript docs

---
title: "Linear Mixed Models: Random Intercept and Slope"
author: "Biostatistics Working Group"

---

# LMM: Intercept & Slope {.unnumbered}

@@

The Linear Mixed Model with a random intercept and slope (LMM:ris) extends the traditional fixed-effect linear regression by incorporating both a subject-specific random intercept and a random slope. This allows each participant to have their own unique intercept and slope values, reflecting individual starting points and rates of change, in addition to the overall mean-level (fixed-effect) trajectory.

In this example, we will utilize the LMM:ris to analyze height trajectories obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand the stability and evolution in height measurements, while factoring in the clustered nature of observations within individuals over time. The LMM:ris facilitates this by concurrently modeling an overarching sample mean trajectory (fixed effect) and individual variations (random effects) in both starting points (intercepts) and growth rates (slopes) around this mean trajectory.

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT","lme4","report","broom","gridExtra","easystats","gtsummary")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(lme4)         # Linear mixed-effects models
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page
library(easystats)    #
library(gtsummary)    #
```
:::

@@

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

@@

@@

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

@@

@@

The code fits a linear mixed model to examine the 'Height' variable across time points ('eventname'). It incorporates both random intercepts and slopes for the time points ('eventname') within each participant ('src_subject_id') to capture individual-level variability in both starting values and rates of change over time. The results of the model are then printed to provide a detailed summary of the fitted model parameters.

**STEP 1: Compute LMM with Random Intercepts and Slopes**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

## Linear Mixed Model with a random intercept and random slope (LMM-ris)
random_intercepts_slopes <- lmer(anthroheightcalc ~ 1 + eventname + (1|src_subject_id), data = df, REML=T)

print(random_intercepts_slopes)

@@

```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Output and reports extending from the LMM-ris analyses
summary(random_intercepts_slopes)
confint(random_intercepts_slopes, level = 0.95, method = "Wald")
report_performance(random_intercepts_slopes)

@@

The code fits a Linear Mixed Model (LMM:ris) to predict children's height based on different time points (denoted as eventname). Individual variability is captured through random intercepts associated with each participant (src_subject_id). The results suggest a significant change in height across the time points, with the model accounting for 89% (r round(conditional_R2, 2)) of the total variation and 37% (r round(marginal_R2, 2)) attributed to the fixed effects alone.

In the model, the linear term for eventname has an estimated effect of r round(model_summary$coefficients["eventname.L", "Estimate"], 2), indicating a significant increase in height across time points. Additionally, there's variability in initial height across participants, reflected by a random intercept standard deviation of r round(random_effects["src_subject_id", "Std.Dev."], 2).

@@

# Assuming your model is named `random_intercepts_slope`
# 1. Extract the random effects
random_effects <- ranef(random_intercepts_slopes)[[1]]

# 2. Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# Plot 1: Histogram
# Plot 1: Histogram
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
  theme_minimal()

print(hist_plot)

@@

# Extract the data frame used in the model
model_data <- random_intercepts_slopes@frame

@@

# Plot
ggplot(df, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +

# Individual estimated height trajectories in faded lines
geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +

# Overall group-mean trajectory in blue with increased thickness
stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
labs(title = "Individual and Group-Mean Height Trajectories",
     x = "Event Name",
     y = "Height") +
  theme_minimal()

@@

The provided code visualizes individual and group-mean height trajectories over different event names. Individual height trajectories for each subject are depicted as faded gray lines, allowing for a clear view of the variability among subjects. In contrast, the overall group-mean trajectory, which represents the average trend across all individuals for each event name, is highlighted in blue. The average height at r mean(df_descriptable$Height) has increased by r mean(df_descriptable$Height_followup - df_descriptable$Height) units from baseline.

@@

The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4). The eventname predictor was statistically significant with a p-value of r format.pval(fixed_effects$Pr(>|t|)[2], digits = 3). The model's overall ability to explain the variance in height was xxxxx, with a conditional R^2 of r report_performance(random_intercepts)$Conditional R2[1], indicating that it accounted for this proportion of the variability in height when considering both fixed and random effects. The marginal R^2 was r report_performance(random_intercepts)$Marginal R2[1], meaning that the fixed effects alone explained this proportion of the variability.

>>>>

---
title: "Linear Mixed Models: Random Intercept"
author: "Biostatistics Working Group"
execute:
  freeze: auto  # re-render only when source changes
  cache: true
---

@@

The LMM:ri is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value zzzzt

In this example, we will use the LMM:ri to analyze trajectories of height obtained across multiple measurement occasions in a sample of youth taking part in the ABCD Study. Our primary aim is to characterize stability and change in height assessments, while accounting for observations that are clustered within youth over time. To do so, we will use the LMM:ri to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.

@@

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

@@

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(lme4)         # Linear mixed-effects models
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

@@

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::

@@

```
#:::

@@

::: blue
**This code creates a descriptives table**
::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

:::
```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

@@

:::

@@

The code fits a linear mixed model to predict the 'Height' variable based on time points ('eventname') and handedness ('Handedness'), while accounting for individual-level variability by including random intercepts for each participant ('src_subject_id'). The results of the model are then printed to provide a summary of the fitted model parameters.

**STEP 1: Compute LMM Model with Random Intercepts**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| cache: true

## Linear Mixed Model with a random intercept (LMM-ri)
model <- lmer(anthroweightcalc ~ 1 + eventname + sex + (1|src_subject_id), data = df, REML=T)

@@

```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

## Output and reports extending from the LMM-ri analyses
#summary(model)
#confint(model, level = 0.95, method = "Wald")
model_performance(model)
model_parameters(model)
model_parameters(model,standardize = "refit")

@@

The code provided executes a linear mixed model (LMM) to predict children's 
height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4), and also takes into account their sex. This model accounts for individual variability in height by including a random intercept for each subject (src_subject_id). The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.

@@

# Extract the random effects
random_effects <- ranef(model)[[1]]

# Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# Plot 1: Histogram
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") + labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
  theme_minimal()

# Plot 2: Residuals vs Fitted Values
resid_plot <- ggplot(NULL, aes(x = fitted_values, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

```{r}
plot(parameters(model)) +
  ggplot2::labs(title = "A Dot-and-Whisker Plot")

plot(model)
theme_set(theme_modern())
result <- estimate_expectation(model, data = "grid")

plot(result)

plot(result, type = "qq")
plot(result, type = "pp")

normality <- check_normality(model, effects = "random")
plot(normality)

heteroscedasticity <- check_heteroscedasticity(model)
plot(heteroscedasticity)

homogeneity <- check_homogeneity(model)
plot(homogeneity)
plot(homogeneity, type = "density")

check_model(model)

check_model(model, panel = FALSE)

distribution <- check_distribution(model)
distribution
plot(distribution)

contrasts <- estimate_contrasts(model, contrast = "sex")
means <- estimate_means(model)
plot(means)
plot(contrasts, means)

params <- model_parameters(model, effects = "fixed")
plot(params)

plot(params, show_labels = TRUE, size_text = 4)

@@

# Extract the data frame used in the model
model_data <- model@frame

@@

# Apply the recoding to the eventname variable
df_subset$eventname <- factor(df_subset$eventname, levels = names(eventname_map), labels = eventname_map)

# Verify the recoding
table(df_subset$eventname)

# Generate the plot
ggplot(df_subset, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +
  # Individual estimated height trajectories in faded lines
  geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +
  # Overall group-mean trajectory in blue with increased thickness
  stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
  labs(title = "Individual and Group-Mean Height Trajectories",
       x = "Event Name",
       y = "Height") +
  theme_minimal()

@@

```{r}
#| echo: false

## Creating variable (names) from output to be used for inline text

library(broom.mixed)
library(easystats)

tidyfit <- tidy(model)
tidyaugment <- augment(model)
tidyglance <- glance(model)

Intercept_estimate <- round(tidyfit[tidyfit$term == "(Intercept)", "estimate"], 3)
Intercept_SE <- round(tidyfit[tidyfit$term == "(Intercept)", "std.error"], 3)

sex_effect <- round(tidyfit[tidyfit$term == "sexMale or other", "estimate"], 3)

# Extract values from the tibble for inline text
Intercept_estimate <- Intercept_estimate$estimate
Intercept_SE <- Intercept_SE$std.error
sex_effect <-sex_effect$estimate

#############

#r eventname_estimate
#r sex_effect
#r conditional_R2
#r conditional_R2*100**%
#r marginal_R2

```{r}
report(model)
```

@@

The estimate of the random intercept was `r Intercept_estimate` (*SE* `r Intercept_SE`). The fixed effects section provides the coefficients for the intercept, eventname, and the sex variable.

The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4) using the event name (eventname) and height (Height) variables. The intercept, corresponding to the reference levels of the predictors is estimated at `r Intercept_estimate` units, and this effect is highly significant. Regarding sex, boys show an increase in height of about `r sex_effect` (*SE* `r Intercept_SE`) units compared to girls.

```{r}
report_performance(model)
```

### 2024-03-06 _posts/_11_Examples_MultivariateLatentGrowthCurveModels.qmd
revised manuscript docs

---
title: "State-Trait Models"
author: "Biostatistics Working Group"
---

@@

------

@@

```{r}
# Define the model
model <- '
  # Trait component
  trait =~ 1 * anthroheightcalc_0 + 1 * anthroheightcalc_1 + 1 * anthroheightcalc_2 + 1 * anthroheightcalc_3

  # State component
  state =~ anthroheightcalc_0 + anthroheightcalc_1 + anthroheightcalc_2 + anthroheightcalc_3

  # Residual variances
  anthroheightcalc_0 ~~ anthroheightcalc_0
  anthroheightcalc_1 ~~ anthroheightcalc_1
  anthroheightcalc_2 ~~ anthroheightcalc_2
  anthroheightcalc_3 ~~ anthroheightcalc_3
'
```

```{r}
# Estimate the State-Trait Model
fit <- sem(model, data = df_wide, missing = "FIML")

>>>>

---
title: "Multivariate Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

```{r}
x <- 7;
x;
```

@@

--------

@@

```{r}
# Specify the multivariate latent growth curve model
model <- '
  # Intercept and slope factors for variable 1
  i_var1 =~ 1*anthroheightcalc_0 + 1*anthroheightcalc_1 + 1*anthroheightcalc_2
  s_var1 =~ 0*anthroheightcalc_0 + 1*anthroheightcalc_1 + 2*anthroheightcalc_2

  # Intercept and slope factors for variable 2
  i_var1 =~ 1*anthroweightcalc_0 + 1*anthroweightcalc_1 + 1*anthroweightcalc_2
  s_var1 =~ 0*anthroweightcalc_0 + 1*anthroweightcalc_1 + 2*anthroweightcalc_2
'
```

```{r}
# Estimate the model
fit <- lavaan::sem(model, data = df_wide, missing = "FIML")

### 2024-03-06 _posts/_8_Examples_AutoregressiveCrosslaggedPanelModels.qmd
revised manuscript docs

---
title: "Multivariate Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

```{r}
x <- 7;
x;
```

# XXXXXXXXXXX {.unnumbered}

@@

XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.

In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. XXXX models facilitate this by xxxxx.

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables

@@

--------

@@

**STEP 2: Compute xxxxx**

```{r}
# Specify the multivariate latent growth curve model
model <- '
  # Intercept and slope factors for variable 1
  i_var1 =~ 1*anthroheightcalc_0 + 1*anthroheightcalc_1 + 1*anthroheightcalc_2
  s_var1 =~ 0*anthroheightcalc_0 + 1*anthroheightcalc_1 + 2*anthroheightcalc_2

  # Intercept and slope factors for variable 2
  i_var1 =~ 1*anthroweightcalc_0 + 1*anthroweightcalc_1 + 1*anthroweightcalc_2
  s_var1 =~ 0*anthroweightcalc_0 + 1*anthroweightcalc_1 + 2*anthroweightcalc_2
'
```

```{r}
# Estimate the model
fit <- lavaan::sem(model, data = df_wide, missing = "FIML")

summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

@@

>>>>

---
title: "Autoregressive Crosslagged Panel Models"
author: "Biostatistics Working Group"
---

# Autoregressive Crosslagged Panel Models {.unnumbered}

@@

Autoregressive Crosslagged Panel (ARCL) Models xxxxxxx. This xxxxxxx.

In this example, we will utilize ARCL models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. ARCL models facilitate this by xxxxx.

@@

# Create a list of required packages
packages_required <- c("tidyverse","rstatix","DT","lavaan")

@@

library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(lavaan)       # Structural equation modeling package

@@

@@

**STEP 2: Compute xxxxx**
```{r}

model <- '
  # Autoregressive paths
  anthroweightcalc_4 ~ a1 * anthroweightcalc_3
  anthroweightcalc_3 ~ a2 * anthroweightcalc_2
  anthroweightcalc_2 ~ a1 * anthroweightcalc_1
  anthroweightcalc_1 ~ a2 * anthroweightcalc_0
  anthroheightcalc_4 ~ a1 * anthroheightcalc_3
  anthroheightcalc_3 ~ a2 * anthroheightcalc_2
  anthroheightcalc_2 ~ a1 * anthroheightcalc_1
  anthroheightcalc_1 ~ a2 * anthroheightcalc_0

# Cross-lagged paths
anthroheightcalc_4 ~ b1 * anthroweightcalc_3
anthroheightcalc_3 ~ b1 * anthroweightcalc_2
anthroheightcalc_2 ~ b1 * anthroweightcalc_1
anthroheightcalc_1 ~ b2 * anthroweightcalc_0

anthroweightcalc_4 ~ b1 * anthroheightcalc_3
anthroweightcalc_3 ~ b1 * anthroheightcalc_2
anthroweightcalc_2 ~ b1 * anthroheightcalc_1
anthroweightcalc_1 ~ b2 * anthroheightcalc_0
'

@@

Fit the ACPM using the lavaan function:

```{r}
# Fit the model to your data (replace 'your_data' with the name of your data frame)
fit <- sem(model, data = df_wide)

# View the results
summary(fit)
